\pkg{cubble}: An \proglang{R} Package for Organizing and Wrangling Multivariate Spatio-temporal Data

### 2023-06-03 2-cubble.Rmd
section 2.5 2.5  Comparison to other spatio-temporal classes

@@

A nested cubble (in the spatial form) has class \code{c("spatial_cubble_df", "cubble_df")} and is built from a rowwise tibble (\code{rowwise_df}). The nested cubble arranges spatial variables in columns and nests temporal variables in a specialised \code{ts} column. The \code{rowwise_df} class is used to work better with the \code{ts} list column to compute per station summary.

@@

<!-- The data \code{climate_flat} is used to illustrate functionality in the \pkg{cubble} package.  The first five rows of the data are shown below: -->

@@

[also talk about the by argument with example]

@@

Some readers may question why a new data structure is needed rather than directly creating a list-column  on the combined data using \code{dplyr::nest_by()}. The reason is that the \code{cubble} object is specifically designed to utilize the spatio-temporal structure when arranging observations in a single object. Moreover, it enables easy pivoting between purely spatial, purely temporal, or unfolded into a combined form.

From the initial \pkg{spacetime} [@spacetime] package. New packages, such as \pkg{stars} [@stars] and \pkg{sftime} [@sftime], have [...]. This section compares and contrast the cubble data structure with other existing alternative, namely \pkg{stars} and \pkg{sftime}.

The `stars` package creates a `stars` object from an array and a dimension object. In many cases, i.e. satellite images, this is convenient, however, there are also many spatio-temporal data are formatted in 2D data frames.

create a dimension object, populate the data into the array

- array in \proglang{R} allows for character, 
- specific datetime class

*make the example simpler*

  - `temporal$var1 %>% matrix(byrow = TRUE, nrow = 2, ncol = 5)`

This happens when different stations have different start date and require other software to reshape the data correctly.

<!-- The temporal data contains two variables are recorded at two stations (`id`) from 2023 March 21st to 2023 March 25th. Notice is that id2 only records from 2023 March 22nd and variable 2 has an `NA` value on  2023 March 23rd at id2. -->

<!-- ```{r echo = FALSE} -->
<!-- library(stars) -->
<!-- spatial <- tibble( -->
<!--   id = c(1,2), -->
<!--   long = c(130, 140), -->
<!--   lat = c(-38, -44), -->
<!--   name = c("station 1", "station 2"))  -->

<!-- spatial -->

<!-- temporal <- tibble( -->
<!--   id = c(rep(1, each = 5), rep(2, each = 4)), -->
<!--   time = c(seq(as.Date("2023-03-21"), as.Date("2023-03-25"), 1), -->
<!--            seq(as.Date("2023-03-22"), as.Date("2023-03-25"), 1)), -->
<!--   var1 = c(3, 6, 8, 2, 5, 7, 9, 14, 5), -->
<!--   var2 = c(1, 3, 4, 2, 6, 7, NA, 5, 8) -->
<!-- ) -->
<!-- temporal -->
<!-- ``` -->

<!-- The following code can be used to create the dimension object: -->

<!-- ```{r echo = TRUE} -->
<!-- spatial_sf <- spatial %>% sf::st_as_sf(coords = c("long", "lat"), crs = 4326)  -->
<!-- station_vec <- st_geometry(spatial_sf) -->
<!-- time_vec <- unique(temporal$time) -->
<!-- var_vec <- spatial_sf$name -->
<!-- d <- st_dimensions(station = station_vec, time = time_vec) -->
<!-- d -->
<!-- ``` -->

<!-- However, the issue comes when casting the variables into an array of the same dimension: -->

<!-- ```{r echo = TRUE, eval = FALSE} -->
<!-- temporal$var1 %>% matrix(byrow = TRUE, nrow = 2, ncol = 5) %>% array(dim = dim(d)) -->
<!-- ``` -->

<!-- Due to the missingness on 2023 March 21st for id2, the array reuses the first value of the variable to fill in the dimension (the missingness at (2,1) is not flagged and 3 is reused to fill in the dimension). -->

<!-- This causes an issue when multiple stations with different recording periods present in the data, commonly found in the weather data.  -->

The approach taken by the \pkg{sftime} package is to combine the spatial and temporal variables together into a joint table. [for sf objects with daily observation, memory efficiency]

Consider the climate data used in the \pkg{cubble} package, \code{climate_aus}. It contains daily precipitation, minimum and maximum temperature for 639 weather stations across Australia in 2020. The converted sftime object is `r library(cubble); #object.size(sftime::st_as_sftime(climate_aus))/object.size(climate_aus)` times the size of the cubble object:

```{r echo = TRUE}
#climate_aus %>% sftime::st_as_sftime() %>% object.size()
climate_aus %>% object.size()
```

>>>>

```{r class, fig.align="center", out.width = "100%", fig.cap = "Illustration of a cubble object in the long form and nested form, along with the associated attributes."}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.007.png"))
```

@@

A nested cubble (in the spatial form) has class \code{c("spatial_cubble_df", "cubble_df")} and is built from a rowwise tibble (\code{rowwise_df}). The nested cubble arranges spatial variables in columns and nests temporal variables in a specialised \code{ts} column. The \code{rowwise_df} class is imposed since it is helpful to manipulate with list column, i.e. compute per station summary of temporal variables, and you may remove it with \code{strip_rowwise()}.

@@

@@

@@

Given the vast spatio-temporal data structure available  in \proglang{R} to represent spatio-temporal data, this section compares and contrast the \pkg{cubble} with other existing alternative, namely \pkg{stars} and \pkg{sftime}.

The \pkg{stars} package [@stars] uses an array structure, as oppose to tibble in the \code{cubble} package, to represent multivariate spatio-temporal data. While both packages support vector and raster data, it is a matter of choice on which structure to use given the application. Analysts working on satellite imageries may prefer the array structure while others originally working with spatio-temporal data in the 2D data frames may find \pkg{cubble} easier to adopt from their existing computing workflow.

The \pkg{sftime} package [@sftime] also builds from a tibble object, with a main focus to handle irregular spatio-temporal data. While \pkg{sftime} can also handle full space-time grids and sparse space-time layouts, \pkg{cubble} can be thought of as simplifying the repetition in spatial variables with two linked tibbles. This simplification gives memory efficiency when data is observed frequent, i.e. daily or subdaily, or the spatial geometry is expensive to repeat, i.e. polygons or multipolygons. Consider the \code{climate_aus} data in the \pkg{cubble} package with `r nrow(cubble::climate_aus)` stations observed daily in a single year 2020. The created \code{sftime} object is `r library(cubble); round(object.size(sftime::st_as_sftime(climate_aus))/object.size(climate_aus), 2)` times larger than the corresponding \code{cubble} object (`r round(object.size(sftime::st_as_sftime(climate_aus))/1e6, 2)` MB vs. `r round(object.size(cubble::climate_aus)/1e6, 2)` MB).

### 2023-06-03 jss/main.tex
section 2.5 2.5  Comparison to other spatio-temporal classes

@@

A nested cubble (in the spatial form) has class \code{c("spatial_cubble_df", "cubble_df")} and is built from a rowwise tibble (\code{rowwise_df}). The nested cubble arranges spatial variables in columns and nests temporal variables in a specialised \code{ts} column. The \code{rowwise_df} class is used to work better with the \code{ts} list column to compute per station summary.

@@

{[}also talk about the by argument with example{]}

@@

Some readers may question why a new data structure is needed rather than directly creating a list-column on the combined data using \code{dplyr::nest_by()}. The reason is that the \code{cubble} object is specifically designed to utilize the spatio-temporal structure when arranging observations in a single object. Moreover, it enables easy pivoting between purely spatial, purely temporal, or unfolded into a combined form.

From the initial \pkg{spacetime} \citep{spacetime} package. New packages, such as \pkg{stars} \citep{stars} and \pkg{sftime} \citep{sftime}, have {[}\ldots{]}. This section compares and contrast the cubble data structure with other existing alternative, namely \pkg{stars} and \pkg{sftime}.

The \texttt{stars} package creates a \texttt{stars} object from an array and a dimension object. In many cases, i.e.~satellite images, this is convenient, however, there are also many spatio-temporal data are formatted in 2D data frames.

create a dimension object, populate the data into the array

\begin{itemize}
\tightlist
\item
  array in \proglang{R} allows for character,
\item
  specific datetime class
\end{itemize}

\emph{make the example simpler}

\begin{itemize}
\tightlist
\item
  \texttt{temporal\$var1\ \%\textgreater{}\%\ matrix(byrow\ =\ TRUE,\ nrow\ =\ 2,\ ncol\ =\ 5)}
\end{itemize}

This happens when different stations have different start date and require other software to reshape the data correctly.

The approach taken by the \pkg{sftime} package is to combine the spatial and temporal variables together into a joint table. {[}for sf objects with daily observation, memory efficiency{]}

Consider the climate data used in the \pkg{cubble} package, \code{climate_aus}. It contains daily precipitation, minimum and maximum temperature for 639 weather stations across Australia in 2020. The converted sftime object is times the size of the cubble object:

\begin{CodeChunk}
\begin{CodeInput}
R> #climate_aus %>% sftime::st_as_sftime() %>% object.size()
R> climate_aus %>% object.size()
\end{CodeInput}
\begin{CodeOutput}
8523040 bytes
\end{CodeOutput}
\end{CodeChunk}

>>>>

{\centering \includegraphics[width=1\linewidth]{../figures/diagram-keynotes/diagram-keynotes.007}

\caption[Illustration of a cubble object in the long form and nested form, along with the associated attributes]{Illustration of a cubble object in the long form and nested form, along with the associated attributes.}\label{fig:class}
\end{figure}
\end{CodeChunk}

@@

A nested cubble (in the spatial form) has class \code{c("spatial_cubble_df", "cubble_df")} and is built from a rowwise tibble (\code{rowwise_df}). The nested cubble arranges spatial variables in columns and nests temporal variables in a specialised \code{ts} column. The \code{rowwise_df} class is imposed since it is helpful to manipulate with list column, i.e.~compute per station summary of temporal variables, and you may remove it with \code{strip_rowwise()}.

@@

@@

Given the vast spatio-temporal data structure available in \proglang{R} to represent spatio-temporal data, this section compares and contrast the \pkg{cubble} with other existing alternative, namely \pkg{stars} and \pkg{sftime}.

The \pkg{stars} package \citep{stars} uses an array structure, as oppose to tibble in the \code{cubble} package, to represent multivariate spatio-temporal data. While both packages support vector and raster data, it is a matter of choice on which structure to use given the application. Analysts working on satellite imageries may prefer the array structure while others originally working with spatio-temporal data in the 2D data frames may find \pkg{cubble} easier to adopt from their existing computing workflow.

The \pkg{sftime} package \citep{sftime} also builds from a tibble object, with a main focus to handle irregular spatio-temporal data. While \pkg{sftime} can also handle full space-time grids and sparse space-time layouts, \pkg{cubble} can be thought of as simplifying the repetition in spatial variables with two linked tibbles. This simplification gives memory efficiency when data is observed frequent, i.e.~daily or subdaily, or the spatial geometry is expensive to repeat, i.e.~polygons or multipolygons. Consider the \code{climate_aus} data in the \pkg{cubble} package with 639 stations observed daily in a single year 2020. The created \code{sftime} object is 13.87 times larger than the corresponding \code{cubble} object (118.24 MB vs.~8.52 MB).
